Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-11-01 analysis/paper/paper.Rmd
Developed discussion, changed "translucent" to "colourless", and other slight updates to text

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 600
)
```

@@

#Fill the NA fields in the munsell_hue column and mark them as translucent samples
Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"

@@

While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these. It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature. In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit* , the latter possibly a spelling mistake], and brecciated quartzite. Those that had no prior material classification registered were sorted into either the quartz or quartzite category. The reason for including *quartz* in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.

@@

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm-1 – 2 500 cm-1 ^[Raman shifts are typically given in wavenumber (cm-1), instead of spectral wavelength (nm)] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.

@@

## X-Ray Fluourescence
XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was set to *mining* mode. Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light). Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions. The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL, or missing values, during PCA. Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, since the former suffers when there are a large number of missing values the latter was used.

@@

The main explanatory power for the visual region can be found in the first component with an explained variance of 98.4 %. In the PCA (fig. \@ref(fig:vis-nir-pca)) there is a clear trend along PC1 with darker materials being grouped together to the right of the plot and light/translucent/white material to the left. Examining the average spectra of these groups (fig. \@ref(fig:vis-nir-spectra)), as expected, the darker material feature higher absorbance of light than the other groups. No other distinctive features can be seen in the visual band, that could potentially be used to distinguish the groups. It is, however, apparent that while the more extreme hues of *dark* and *white* material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups.

A number of artefacts also breaks the observed trend in the PCA, with dark/black (N1-N2) samples ending up closer to the light and translucent group in the PCA. There are also instances where the measurements on different sides of an artefact end up in different groups in the PCA. In examination of the non-averaged data artefact 403 and 502 both have a sample taken on one side in the *dark* group and the other in the *greyish* middle. This is most likely due to the heterogeneous character of the material and natural variations on the surface possibly related to weathering. The *greyish* middle group in the PCA is otherwise largely made up of *quartz coloured* opaque material with a beige/grey tint.

@@

```{r vis-nir-pca, fig.width=20, fig.height=8, fig.cap="PCA score plot of PC1 and PC2 for the 400 nm – 750 nm band of the spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant. The strong trend along PC1 also makes it difficult to separate out the muddled grayscale in the middle."}

@@

#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
basic_plot1 <-
  fviz_pca_ind(visnir.pca, axes = c(1,2), label="none")

#prepare labels for PCs
vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 400 - 750 nm") +
  labs(x = vis.pc1lab,
       y = vis.pc2lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", 
                             title.hjust = 0.5)) +
  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
        legend.title = element_text(size = 12, face = "bold", colour = "black"),
        legend.background = element_rect(linetype = "solid", color = "black"),
        legend.position = "bottom")
```

@@

vis.translucent <- Points.nir %>% 
  filter(hue == "Colourless") %>% 
  select(c(1:28),c(79:429))

spec.vt <- 
  as_spectra(vis.translucent, name_idx = 1, meta_idxs = 2:28)

@@

The PCA shows a strong trend along PC 1 that seems to correlate with the colour of the material(fig. \@ref(fig:nir-pca)). Darker quartzite cluster to the right with white and translucent samples grouping to the opposite side. While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the *brecciated quartz*, seem to cluster towards the top-left. A small set of 3 black samples at the top of the model separate themselves clearly from the rest of the data points. These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.

@@

```{r nir-spectra, fig.width=20, fig.height=15, fig.cap="Average NIR spectra by hue. Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/translucent samples."}

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:28),c(679:2179))

#create a spectra object for the dark material, and then repeat for light, white and translucent groups
spec.d <- 
  as_spectra(dark, name_idx = 1, meta_idxs = 2:28)

light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:28),c(679:2179))

@@

translucent <- Points.nir %>% 
  filter(hue == "Colourless") %>% 
  select(c(1:28),c(679:2179))

spec.t <- 
  as_spectra(translucent, name_idx = 1, meta_idxs = 2:28)

white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:28),c(679:2179))

@@

## Raman
There are two distinctive features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first are three peaks in the 100 cm-1 - 500 cm-1 band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm-1 [@gillet_etal90]. These peaks are present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm-1 region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm-1 and 795 cm-1. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group. Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D/G peaks. PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data. Interestingly, the loading for PC 4 does catch the Al-O-Si peaks around the  700 cm-1 - 800 cm-1 band, which helps in separating out those samples where this feature is present. The measurements with mainly quartz peaks gravitates towards the negative side of the PC 1, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the *quartz* group.

@@

```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue. The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the dark group"}

@@

r.translucent <- Points.raman %>% 
  filter(hue == "Colourless")

spec.rt <- 
  as_spectra(r.translucent, name_idx = 1, meta_idxs = 2:29)

@@

#Plot the mean spectrum for each spectra group above with text annotations for the peaks in the first dark spectra
plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
     xlab = "Wavenumber (cm-1)", main = "Average - Dark")
text(x=120, y=2900, labels="SiO", font=2, cex = 1.5)
text(x=200, y=4300, labels="SiO", font=2, cex = 1.5)
text(x=460, y=9250, labels="SiO", font=2, cex = 1.5)
text(x=1285, y=6300, labels="D", font=2, cex = 2.0)
text(x=1600, y=4000, labels="G", font=2, cex = 2.0)

plot(mean(spec.rl), lwd = 1.2, col = "red", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
     xlab = "Wavenumber (cm-1)", main = "Average - Light")

plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
     xlab = "Wavenumber (cm-1)", main = "Average - Colourless")

plot(mean(spec.rw), lwd = 1.2, col = "blue", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
     xlab = "Wavenumber (cm-1)", main = "Average - White")
```

@@

```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra. The main explanatory power can be found in the 1^st^ component, which groups the data based on the quartz and D/G bands. As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA. The 3^rd^ component does, however, also separate out the few samples that feature peaks at 700-800 cm-1 that may result from Al, Si and OH vibrations."}

@@

The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 84.7%. The loadings of the three components can be seen in figure \@ref(fig:xrf-loadings). In the score plot of PC 1 and PC 2, two large clusters can be seen (fig. \@ref(fig:xrf-score-plot-p1-p2)). Interestingly, the smaller cluster on the left side in the model includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content. It is not ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that drives this. The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom. This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite ([@howard05] p.711). The *brecciated quartz* samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up towards closer to the “high ^14^Si” cluster. The white (N9) material is somewhat evenly distributed within the right cluster.

@@

One almost translucent sample also ends up in the "high ^14^Si" cluster among the darker material. Upon closer inspection this sample is small, ~ 3 by 1 mm, and also feature inclusions of dark particles. Both of these factors may have contributed to its placement in the model.

@@

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
#Prepare data in long format for violin plots
Points_long <-
  Points.xrf %>%
  select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
  gather(key = "Element", value = "Value")

@@

@@

#Violin plot of elemental content, trimmed, with fill based on elemental groups
Points_long %>%
  drop_na(Value) %>%
    ggplot(aes(x=Element, y=Value, fill = Group)) +
    geom_violin(trim = TRUE) +
    facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
    scale_fill_manual(values=loading.colors) +
    theme_bw() +
    theme(
      legend.title = element_text(size = 12, face = "bold", colour = "black"),
      strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.x = element_blank(),
      plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
      axis.title.y=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank())
```

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

@@

#Fill the NA fields in the munsell_hue column and mark them as colourless samples
Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"

@@

While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these. It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature. In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz and brecciated quartzite^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit*]. Those that had no prior material classification registered were sorted into either the quartz or quartzite category. The reason for including *quartz* in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.

@@

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^ ^[Raman shifts are typically given in wavenumber (cm^-1^), instead of spectral wavelength (nm)] with a resolution of 9.5 cm^-1^. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.

@@

## X-Ray Fluourescence
XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was set to *mining* mode. Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light). Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions. The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL, or missing values, during PCA. Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

@@

The main explanatory power for the visual region can be found in the first component with an explained variance of 98.4 %. In the PCA (fig. \@ref(fig:vis-nir-pca)) there is a clear trend along PC1 with darker materials being grouped together to the right of the plot and light/colourless/white material to the left. Examining the average spectra of these groups, as expected, the darker material feature higher absorbance of light than the other groups. No other distinctive features can be seen in the visual band, that could potentially be used to distinguish the groups. It is, however, apparent that while the more extreme hues of *dark* and *white* material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups. There are also a number of artefacts with a dark hue that gravitate closer to the lighter material in the PCA. While the small size of these artefacts could be an explanation for the placement of these within the model, there are plenty of similarly sized artefacts that do not follow the same trend.

There are also instances where the measurements on different sides of an artefact end up in different groups in the PCA. In examination of the non-averaged data artefact 403 and 502 both have a sample taken on one side in the *dark* group and the other in the *greyish* middle. This is most likely due to the heterogeneous character of the material and natural variations on the surface possibly related to weathering. The *greyish* middle group in the PCA is otherwise largely made up of opaque material with a beige/grey tint.

@@

```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plot of PC1 and PC2 for the 400 nm – 750 nm band of the spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant. The strong trend along PC1 also makes it difficult to separate out the muddled grayscale in the middle."}

@@

#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
basic_plot1 <-
  fviz_pca_ind(visnir.pca, axes = c(1,2), label="none")
basic_plot2 <-
  fviz_pca_ind(visnir.pca, axes = c(1,3), label="none")

#prepare labels for PCs
vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
vis.pc3var <- round(summary(visnir.pca)$importance[2,3]*100, digits=1)
vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")
vis.pc3lab <- paste0("PC3 (",as.character(vis.pc3var),"%)")

#VISNIR LOADING LINES
#par(mfrow=c(3,1))
#sapply(1:3, function(comp){
#  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,6))})

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
visnir.p1 <-
  ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 400 - 750 nm") +
  labs(x = vis.pc1lab,
       y = vis.pc2lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", 
                             title.hjust = 0.5)) +
  theme(legend.position = "none")

visnir.p2 <-
  ggplot(cbind(basic_plot2$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 400 - 750 nm") +
  labs(x = vis.pc1lab,
       y = vis.pc3lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", 
                             title.hjust = 0.5)) +
  theme(legend.position = "none")

comb1 <-
  visnir.p1 + visnir.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
                              axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
                              legend.title = element_text(size = 12, face = "bold", colour = "black"),
                              legend.background = element_rect(linetype = "solid", color = "black"),
                              legend.position = "bottom")

comb1 +
  plot_layout(nrow = 2, guides = "collect")
```

@@

vis.colourless <- Points.nir %>% 
  filter(hue == "Colourless") %>% 
  select(c(1:28),c(79:429))

spec.vt <- 
  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)

@@

The PCA shows a strong trend along PC 1 that seems to correlate with the colour of the material(fig. \@ref(fig:nir-pca)). Darker quartzite cluster to the right with white and colourless samples grouping to the opposite side. While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the *brecciated quartz*, seem to cluster towards the top-left. A small set of 3 black samples at the top of the model separate themselves clearly from the rest of the data points. These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.

@@

```{r nir-spectra, fig.width=20, fig.height=15, fig.cap="Average NIR spectra by hue. Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/colourless samples."}

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:28),c(680:2179))

#create a spectra object for the dark material, and then repeat for light, white and colourless groups
spec.d <- 
  as_spectra(dark, name_idx = 1, meta_idxs = 2:28)

light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:28),c(680:2179))

@@

colourless <- Points.nir %>% 
  filter(hue == "Colourless") %>% 
  select(c(1:28),c(680:2179))

spec.t <- 
  as_spectra(colourless, name_idx = 1, meta_idxs = 2:28)

white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:28),c(680:2179))

@@

## Raman
There are two distinctive features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first are three peaks in the 100 cm^-1^ - 500 cm^-1^ band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm^-1^ [@gillet_etal90]. These peaks are present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm^-1^ and 795 cm^-1^. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band which has been related to Al-O-Si vibrations for the Kaolin group. Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D/G peaks. PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data. Interestingly, the loading for PC 4 does catch the Al-O-Si peaks around the  700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present. The measurements with mainly quartz peaks gravitates towards the negative side of the PC 1, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the *quartz* group.

@@

```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue. The *quartz peak* at 460 cm^-1^ is present in all four groups, but the *graphite peaks* at 1 287  cm^-1^ (D band) and 1 598 cm^-1^ (G band) mainly appear in the dark group"}

@@

r.colourless <- Points.raman %>% 
  filter(hue == "Colourless")

spec.rt <- 
  as_spectra(r.colourless, name_idx = 1, meta_idxs = 2:29)

@@

#Plot the mean spectrum for each spectra group above with text annotations for the peaks in the first dark spectra
plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
     xlab = "Wavenumber (cm^-1^)", main = "Average - Dark")
text(x=120, y=2900, labels="SiO", font=2, cex = 1.5)
text(x=200, y=4300, labels="SiO", font=2, cex = 1.5)
text(x=460, y=9250, labels="SiO", font=2, cex = 1.5)
text(x=1285, y=6300, labels="D", font=2, cex = 2.0)
text(x=1600, y=4000, labels="G", font=2, cex = 2.0)

plot(mean(spec.rl), lwd = 1.2, col = "red", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
     xlab = "Wavenumber (cm^-1^)", main = "Average - Light")

plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
     xlab = "Wavenumber (cm^-1^)", main = "Average - Colourless")

plot(mean(spec.rw), lwd = 1.2, col = "blue", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
     xlab = "Wavenumber (cm^-1^)", main = "Average - White")
```

@@

```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra. The main explanatory power can be found in the 1^st^ component, which groups the data based on the quartz and D/G bands. As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA. The 3^rd^ component does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ that may result from Al, Si and OH vibrations."}

@@

The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 81.3%. The loadings can be be seen in figure \@ref(fig:xrf-loadings). In the score plot of PC 1 and PC 2 two large clusters can be seen (fig. \@ref(fig:xrf-score-plot-p1-p2)). Interestingly, the smaller cluster on the left side in the model includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content. It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr. The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom. This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite [@howard05 p.711]. The *brecciated quartz* samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up closer towards the “high ^14^Si” cluster. The white (N9) material is fairly evenly distributed within the right cluster.

@@

One of the *colourless* samples also ends up in the "high ^14^Si" cluster among the darker material. Upon closer inspection this sample is small, ~ 3 by 1 mm, and also feature inclusions of dark particles. Both of these factors may have contributed to its placement in the model.

@@

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
#Prepare data in long format for violin plots
Points_long <-
  drop_na(Points.xrf %>%
    select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
    gather(key = "Element", value = "Value"))

@@

#Calculate and add value equal to 95% of the value range for each element, this is in preparation for 
#text annotations in final plot where a text will be positioned at this values height in each plot
Points_long <- 
  Points_long %>% 
  group_by(Element) %>% 
    mutate(ypos = min(Value)+0.95*diff(range(Value)))

@@

#Prepare data frame with elements for annotating facets in final plot
na_values <- 
  data.frame("Element" = unique(Points_long$Element))

#Calculate and add number of missing values for each element             
na_values <- 
  na_values %>% 
    mutate(na_values = c(
      paste("n = ", length(which(!is.na(Points.xrf$`12Mg`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`13Al`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`14Si`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`15P`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`16S`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`19K`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`20Ca`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`23V`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`25Mn`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`26Fe`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`30Zn`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`38Sr`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`39Y`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`40Zr`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`56Ba`)))),
      paste("n = ", length(which(!is.na(Points.xrf$`81Ti`)))))
    )

#Merge na_values and Points_long data frames in order to get the y position that the text annotation should be displayed
#in for each element facet in final plot
na_values <- 
  merge(na_values, unique(Points_long[,c("Element", "ypos")]), by = "Element", all.x = TRUE)
